---
title: "miRNA pipeline: differential expression analysis"
author: "RWH"
date: "`r format(Sys.time(), '%d %B %Y')`"

output: 
  html_document:
    theme: yeti 
    highlight: pygments 
    anchor_sections: FALSE
---

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = TRUE) 

library(tidyverse)
library(here)
library(edgeR)

dir_scripts <- here(dir_root, "scripts")
dir_save <- here(dir_data, "RMD_output")

source(here(dir_scripts, "plotting_themes.R"))
source(here(dir_scripts, "plotting.R"))
source(here(dir_scripts, "analysis.R"))

```

```{r set_parameters}

# DEA analysis
fdr_threshold <- 0.05   # 0.05
fc_threshold <- 1.5     # 2

```

```{r load_dgl}

dea_objects <- readRDS(file = here(dir_save, "dea_objects.rds"))

dgl <- dea_objects$dgl
df_samples <- dea_objects$df_samples
contrast_name <- dea_objects$contrast_name

```


<br>
<br>

## Analysis parameters  

In the differential expression analysis, genes that are `r fc_threshold`-fold differentially expressed with an FDR < `r fdr_threshold` are considered significant.  

<br>
<br>


## Differential expression analysis

Construct a simple two-group model - or add in additional variables in a fixed-effect model.  

```{r DEA}

design <- model.matrix(~0+group, data = dgl$samples)
dgl <- estimateDisp(dgl, design)

contrast_matrix <- makeContrasts(contrasts=contrast_name, levels=design)
fit <- glmQLFit(dgl, design)

de <- glmQLFTest(fit, contrast = contrast_matrix)

```


Top 20 most differentially expressed genes are:  

```{r top_20_dea }

top_genes <- topTags(de, n=20, adjust.method = "BH") # Benjamini & Hochberg FDR)
top_genes$table %>% knitr::kable()

df_dea_results <- topTags(de, n=nrow(dgl), sort.by='none', adjust.method = "BH")$table

```


```{r volcano_plot}

df_dea_results %>% 
  df_prep_volcano(fc_threshold = fc_threshold, fdr_threshold = fdr_threshold) %>% 
  volcano_plot_RWH() + sc_volcano

DE_genes <- rownames(df_dea_results)[abs(df_dea_results$logFC) >= log2(fc_threshold) &
                                         df_dea_results$FDR <= fdr_threshold ] 

```

`r length(DE_genes)` genes are differentially expressed with a fold-change of at least `r fc_threshold` and an FDR below `r fdr_threshold`.  


<br>
<br>

### Top genes

Expression plots for top 8 differentially expressed genes:  

```{r plot_top_genes}

plot_gene_dotplot_RWH_many(rownames(top_genes$table)[1:4], dgl) + sf_miR_groups

# plot_gene_dotplot_RWH(rownames(top_genes$table)[1], dgl) + sf_miR_groups
# plot_gene_dotplot_RWH(rownames(top_genes$table)[2], dgl) + sf_miR_groups
# plot_gene_dotplot_RWH(rownames(top_genes$table)[3], dgl) + sf_miR_groups
# plot_gene_dotplot_RWH(rownames(top_genes$table)[4], dgl) + sf_miR_groups


```

```{r save_dea_genes}

df_dea_results %>% write_csv(here(dir_save, "dea.csv"))

```